Investigating the Impact of Macroeconomic Indicators on the FTSE 250 Index Using Machine Learning Methods

### 2024-08-14 tex/paper.tex
update
lit review

A key task which economic policy makers and investors undertake is to understand and predict how economic changes influence stock market behaviour, which is critical for making informed policy and investment decisions. This modelling of the impact of macroeconomic indicators on stock markets is typically achieved by a variety of methods: statistical analysis, econometric modelling, scenario planning, and advanced machine learning techniques. However, it is fraught with challenges that require careful consideration, advanced techniques, and a deep understanding of both economics and financial markets. In particular, there is a challenge to find a subset of macroeconomic indicators which can most accurately capture stock market dynamics.

@@

There are a number of works which investigate the impact of 
the selected macroeconomic variables
on the stock market. However, as of yet the effect of the entirety of the selected 
macroeconomic variables has not been studied on the FTSE250.

The FTSE250 has historically been considered a more closer representation of the domestic economy than the FTSE100. Roughly $50\%$ of the earnings of the FTSE250 are from overseas, similar to the S\&P 500. In contrast, nearly $80\%$ of the FTSE$100$'s earnings are from overseas. 
This is reflected in the constitution of the respective indices, with the 
FTSE250 historically resembling the structure of the UK economy more closely than the FTSE100 \citep{ftse250history}.
In spite of this, the FTSE250 has seldom been utilised as the preferred indicator of the UK stock market in the literature.

@@

\subsection{Forecasting}

@@

RBF-NNs have been used 
considerably for stock market forecasting. \citeboth{cao2004} 
used an RBF-NN based on the Optimal Partition Algorithm (OPA) to 
forecast S$\&$P 500 prices. \citeboth{dass2019} use an RBF-NN, with the Back-propagation 
Algorithm (BPA) used instead to tune parameters; their results demonstrate 
that the RBF-NN outperformed a standard feed-forward neural network. 
\citeboth{ji2014} also found that the RBF-NN was able to predict the value 
of the Shanghai Composite Index on a 45 day time-horizon, with a mean prediction error of 
1.4387$\%$. Recently, \citeboth{abotaleb2024}
used quarterly data from 2001 to 2023 to model the influence of economic 
policies on the quarterly prices of various stock markets including the 
FTSE100 using an RBF-NN, finding that the model 
achieved relative forecast error rates as low as 6.8$\%$ and an error of 
27.2$\%$ in the case of the FTSE100. Thus, it is concluded that the RBF-NN is a 
competent tool in the sphere of financial time-series forecasting, and 
is able to operate effectively within the constraints placed by our 
problem domain and data. However, with the exception of \citeboth{abotaleb2024}, who used 
quarterly data, the effectiveness of RBF-NNs in stock market forecasting using 
low-granularity economic time-series has not been investigated and presents an area requiring further 
research.

@@

    \For{each month $m$}{
        \If{$\forall d \in D_m \; \text{(interest rate unchanged)}$}{
            \textbf{Set:} $\text{Rate}_m \leftarrow \text{Rate}_{m-1}$\;
        }
        \Else{
            \For{each day $d_j$ with a rate change}{
                \textbf{Compute:} 
                $W_{int} = \sum_{i_k \in I_m} i_k \frac{d_{j_k}}{x}$\;
                where $I_m = \{i_1, i_2, \ldots, i_n\}$ is the set of interest rates on days with changes, 
                and $d_{j_k}$ is the number of days for each rate $i_k$.
            }
            \textbf{Set:} $\text{Rate}_m \leftarrow W_{int}$\;
        }
    }
\end{algorithm}

>>>>

A key task which economic policy makers and investors undertake is to understand and predict how economic changes influence stock market behaviour, which is critical for making informed policy and investment decisions. This modelling of the impact of macroeconomic indicators on stock markets is typically achieved by a variety of methods: statistical analysis, econometric modelling, scenario planning, and advanced machine learning techniques. However, it is fraught with challenges that require careful consideration, advanced techniques, and a deep understanding of both economics and financial markets. In particular, there is a challenge to find a subset of macroeconomic indicators which can most accurately capture stock market dynamics. 
In addition, macroeconomic indicator data is of such low-granularity that it precludes the use of certain deep learning neural network methods for prediction purposes.

@@

There are a number of works which investigate the impact of 
the selected macroeconomic variables
on the stock market.

@@

The FTSE250 has historically been considered a more closer representation of the domestic economy than the FTSE100. Roughly $50\%$ of the earnings of the FTSE250 are from overseas, similar to the S\&P 500. In contrast, nearly $80\%$ of the FTSE$100$'s earnings are from overseas. 
This is reflected in the constitution of the respective indices, with the 
FTSE250 historically resembling the structure of the UK economy more closely than the FTSE100 \citep{ftse250history}.

In spite of this, the FTSE250 has seldom been utilised as the preferred indicator of the UK stock market in the literature.
With the exception of \citeboth{smit2023longrun}, \citeboth{khatri2024impact} and \citeboth{engstrom2006impact} who investigate the effect of variable changes on the FTSE All Share which aggregates the 
FTSE100, FTSE250 and FTSE SmallCap indices, as of yet, the impact of the four selected variables on the FTSE250 has not been empirically studied and 
warrants further investigation.

\subsection{Stock Market Forecasting}

@@

RBF-NNs have been used 
considerably for stock market forecasting. \citeboth{cao2004} 
used an RBF-NN based on the Optimal Partition Algorithm (OPA) to 
forecast S$\&$P 500 prices. \citeboth{dass2019} use an RBF-NN, with the Back-propagation 
Algorithm (BPA) used instead to tune parameters; their results demonstrate 
that the RBF-NN outperformed a standard feed-forward neural network. 
\citeboth{ji2014} also found that the RBF-NN was able to predict the value 
of the Shanghai Composite Index on a 45 day time-horizon, with a mean prediction error of 
1.4387$\%$. Recently, \citeboth{abotaleb2024}
used quarterly data from 2001 to 2023 to model the influence of economic 
policies on the quarterly prices of various stock markets including the 
FTSE100 using an RBF-NN, finding that the model 
achieved relative forecast error rates as low as 6.8$\%$ and an error of 
27.2$\%$ in the case of the FTSE100. Thus, it is concluded that the RBF-NN is a 
competent tool in the sphere of financial time-series forecasting, and 
is able to operate effectively within the constraints placed by our 
problem domain and data.

However, with the exception of \citeboth{abotaleb2024}, who used 
quarterly data and the economic uncertainty index, the effectiveness of RBF-NNs in stock market forecasting using 
major macroeconomic indicators and low-granularity time-series has not been investigated and presents an area for further 
research.

@@

    \For{each month $m$}{
        \If{$\forall d \in D_m \; \text{(interest rate unchanged)}$}{
            \textbf{Set:} $\text{Rate}_m \leftarrow \text{Rate}_{m-1}$\;
        }
        \Else{
            \For{each day $d_j$ with a rate change}{
                \textbf{Compute:} 
                $W_{int} = \sum_{i_k \in I_m} \frac{i_k  d_{j_k}}{x}$\;
                where $I_m = \{i_1, i_2, \ldots, i_n\}$ is the set of interest rates on days with changes, 
                and $d_{j_k} \leq x$ is the number of days into the month each rate $i_k$ occurs on.
            }
            \textbf{Set:} $\text{Rate}_m \leftarrow W_{int}$\;
        }
    }
\end{algorithm}
